An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging

### 2019-10-31 HowtoHowtoCapon/submission/submit2/howto_howto_capon.tex
2019-10-30

\author{Jo~Inge~Buskenes, %
        Jon~Petter~\AA{}sen, %
        Carl-Inge~Colombo~Nilsen, %
        Andreas~Austeng%
\thanks{J. I. Buskenes, C.-I. C. Nilsen and A. Austeng are with the Department of Informatics, University of Oslo, Norway.}%
\thanks{J. P. \AA{}sen is with MI-Lab, Norwegian University of Science and Technology, Trondheim, Norway.}% <-this % stops a space
% \thanks{Manuscript received April 19, 2005; revised January 11, 2007.}
}

@@

In this study we show that for active sonar systems up to 32 channels, the computation time can be significantly reduced by performing arithmetic optimizations, and by implementing the MVDR on a Graphics Processing Unit (GPU). We point out important hardware limitations for these devices, and assess the design in terms of how efficiently it is able to use the GPU's resources. On a quad-core Intel Xeon system with a high-end Nvidia GPU, our GPU implementation renders more than 1\;Mpx/s. This represents a speedup of more than two orders of magnitude over our initial single threaded implementation in C. This enables real-time processing of sonar data, and makes the MVDR a viable alternative to conventional methods in practical systems.
% If the HISAS1030 was attached to a platform traveling at 1.8m/s, creating full-coverage sectorscan images in real-time would be possible with this performance.

% We benchmarked optimized implementations of the MV beamformer in Matlab and C, running on a quad-core Intel Xeon E5620, and in CUDA running on a high-end nVidia card. The beamformers operated on data from the 32x4 element Kongsberg Martime HISAS1030 sonar, using a conservative set of robustification parameters. 
%The CPU would then be free to take on the rest of the beamforming as well.
\end{abstract}

@@

% Robustification:
% R estimate accuracy vs. dynamic trackability: (1) Trade observation time for bandwidth (2) Partial adaptiveness (less degrees of freedom than channels) (3) Quadratic constraint on sensor weights to limit the norm to some preset value.
% -  
To ensure MVDR robustness, the litterature suggests combining means such as temporal and spatial averaging, and regularization~\cite{VanTrees2002,Kailath1985}. The complexity issue, on the other hand, can be handled by introducing well-founded approximations. For instance, some studies assume data stationarity which allows the formation of a Toeplitz-structured covariance matrix that is simpler to invert~\cite{Asl2012a,Jakobsson2000}. Other studies perform MVDR beamforming in beamspace (the spatial frequency domain), which can be considered sparse due to the limited angular extent of the received beam~\cite{VanTrees2002,Nilsen2009}. Exploiting this can lead to considerable performance improvements, especially in high channel count systems.

@@

In this work, we show that such approximations may not be necessary for sonar systems up to 32 channels, since here the computation time can be reduced by 1-2 orders of magnitude by implementing MVDR on a Graphics Processing Unit (GPU). Similar work may be found in e.g. medical ultrasound imaging, where Chen \emph{et al.} have demonstrated a GPU implementation of the MVDR that operates on real valued data~\cite{Chen2011,Chen2011a}. While this implementation is fast, the real valued data constraint does not allow the creation of asymmetric and shifted responses, and hence prevents the MVDR's potential to be fully reached. Our implementation is not restricted in this manner. In related work, we investigated using the same MVDR GPU implementation in cardiac ultrasound imaging~\cite{Asen2012,Asen2013}, and in active sonar imaging we have compared the performance of this implementation to comparable adaptive methods~\cite{Buskenes2013a}.

@@

Various adaptive beamformers target this limitation by allowing the weights to change for each pixel to better fit the dynamic nature of the incoming wavefield. In other words, they attempt to use the \emph{a priori} information present in the data to improve image quality. The MVDR beamformer is one such method. It finds the set of complex weights that minimizes the beamformer's expected output power, while ensuring unity gain in the look direction~\cite{Capon1969}. This is a convex optimization problem that can be solved using Lagrange multipliers to yield the solution:
\begin{gather}
\vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
\end{gather}
where $\1$ is a row vector of ones that represents broadside steering, and $\R=E\{\x[n]\x\H[n]\} \in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array. Since $\R$ is unknown, we estimate it by computing a sample covariance matrix $\eR$. In this computation we will perform some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics~\cite{Synnevag2009}, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Cox1987,Maksym1979}. Combined, these steps will also ensure a numerically well conditioned $\eR$.

To form $\eR$ we will first compute an intermediate sample covariance matrix $\breve{\R}$, which we form using temporal and spatial averaging. To do this we need to segment our array into subarrays. If we let $x_l[n]$ represent the data vector from subarray $l$,
\begin{gather}
\x_l[n] = \bmat{x_l[n] & x_{l+1}[n] & \dots & x_{l+L-1}[n]}\T,
\end{gather}
then $\breve{\R}$ can be calculated as:
\begin{gather}
\breve{\R}[n] =  \frac{1}{N_K N_L} \sumb{l=0}{M-L}\sumb{n'=n-K}{n+K} \x_l[n]\x_l\H[n] \in\mathbb{C}^{L,L},\label{spatialR}
\end{gather}
where $N_K = 2K+1$ is the number of temporal samples to perform averaging over, and $N_L = M-L+1$ is the number of subarrays.

@@

We decided to investigate the feasibility of running MVDR on a GPU by mapping it to a Nvidia GeForce Quadro 6000. This is a high-end Compute Unified Device Architecture (CUDA) enabled GPU based on Nvidia's Fermi architecture. The code was written in Nvidia's ``C for CUDA'' framework. We believe that GPUs from AMD, or the cross-platform OpenCL framework from the Khronos group, are equally attractive alternatives, but comparing them is beyond the scope of this study.

@@

As observed in Fig. \ref{mvdr_complexity}, a direct computation of $\eR$ by implementing (\ref{spatialR}) and (\ref{finalR}) is the greatest computational burden. We target this issue by first performing \emph{arithmetic reductions}, then aim to get as close to the peak \emph{arithmetic throughput} on the GPU as possible.

\subsubsection{Arithmetic Reductions}

In Fig. \ref{mvdr_build_R} we illustrate how to reduce the arithmetic complexity of building $\eR$ in a system with $M=5$ sensors, a subarray size of $L=3$ and temporal averaging is set to $K=1$. Here the spatial sum is carried out before the temporal sum. To reduce arithmetic operations we may:
%
\begin{enumerate}
\item exploit the fact that $\eR$ is Hermitian positive semi-definite to compute only one half of it,
\item avoid redundant multiplications, and
\item implement the spatial sum iteratively by sliding along the diagonals of $\eR$.
\end{enumerate}
%
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske5.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/buildR-breakdown.eps}
\fi%
\caption{Arithmetic optimization of computing $\eR$: Relative reduction in arithmetic complexity compared to the initial implementation shown in Fig. \ref{mvdr_complexity} (higher is better) . Note how the arithmetic count is reduced by a factor 4-10 in the memory optimized case, and by a factor 6-22 in the instruction optimized case.}\label{mvdr_complexity_optimized}
\end{figure}

To study the effect of these optimizations we altered the complexity formulas to take them into account. Exploiting symmetry and performing iterative summing (step 1 and 3) is always desirable, but avoiding redundant multiplications (step 2) comes at the cost of increased memory consumption. The effect of adding all the optimizations, or all but saving multiplications, is shown in Fig. \ref{mvdr_complexity_optimized}. Here the complexity curves are relative to the reference implementation in Fig. \ref{mvdr_complexity}. Both solutions reduce the complexity considerably, and recomputing the multiplications does not make that much of a difference. We selected the memory minimized version since our algorithm is severely memory bound, as we will see next.

@@

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske6.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/mvdr_implementation.eps}
\fi%
\caption{MVDR implementated on a GPU. We do this in 3 steps, where each step process the full image before moving on to the next step. \emph{Step~1}: The sample covariance matrix $\eR$ is formed by threads running along its diagonals. This allows spatial averaging to be implemented in a computationally efficient manner and minimizes inter-thread communication. \emph{Step~2}: $\eRi\1$ is computed using the heavily optimized batched linear equation solver from Nvidia. \emph{Step~3}: The beamformer output $z$ is computed in a straight forward fashion by $L$ threads that first sum the subarrays and then apply the MVDR weighting function. A single thread finally sum all the channels up.}\label{mvdr_implementation}
\end{figure}

@@

When we compute $\eR$, we very rarely perform more than 1-3 floating point operations for every float read or written to memory. Unfortunately, the GPU prefers kernels to be more computationally intensive than this. This can be inferred from Table \ref{throughputs}, where the peak bandwidth of the three types of GPU memory are compared to the peak arithmetic throughput (see appendix \ref{throughput} for derivations). Global memory (RAM), in which all data must reside at some point, is at best only able to move one float for every 30 floats processed by the CUDA cores. This is why the usage of global memory should be minimized. Shared memory, on the other hand, is a very fast level 1 cache that is shared by all threads on an SM. At peak performance it is almost able to keep up with the computing units, but to obtain this performance we must make sure that:
\begin{enumerate}
\item the data needed to compute a pixel can fit in shared memory, while
\item the access patterns we use promotes maximum bandwidth.
\end{enumerate}
 %Further alleviation is also possible by exposing instruction level parallelism within a thread, as this promotes the use of registers with a bandwidth at least 6 times higher than that of shared memory~\cite{Vasilyy}\todo{so why don't we? rephrase}.

@@

In Fig. \ref{mvdr_implementation} we illustrate how we got around these challenges. First we make each SM compute entire range lines of pixels, then load the subset of $\x$ that all these pixels depend upon into shared memory. This lets us perform temporal averaging without having to read additional channel data.

\ifPeerReview\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/plot_holmengraa_L16_Navg1.eps}
\caption{HISAS sidescan sonar (SSS) image of the shipwreck Holmengraa that lies on a slanted seabed at 77\;m depth outside of Horten, Norway. The image was processed with $M=32$, $L=16$, $K=1$ and $d=1\%$. Prior to display the image was linearly upinterpolated by a factor 2 in azimuth making its total size 1.46\;Mpx. Note how MVDR improves edge definition and reduces noise in shadow regions.}\label{holmengraa}
\end{figure}
\fi%

Second, we assign $L$ threads per pixel to traverse the diagonals of $\eR$. On the top row a full computation is carried out, then that row is saved back to global memory following a collective access pattern that maximizes global memory bandwidth. For subsequent rows the threads move along the diagonals while performing iterative summations; the result from the previous element on the diagonal is updated by adding and removing the correlation coefficients that enters and exits the sum, respectively. To minimize memory consumption, we compute the coefficients again every time we need them. When a thread has finished up a diagonal, we have them wrap around to compute one of the diagonals in the lower triangular of $\eR$. Since $\eR$ is conjugate symmetric, the values in the leftmost column is obtained by a complex conjugate copy of the relevant value in the first row. Combined these steps balance the load evenly on all threads, is almost completely free of arithmetic redundancy, and consumes less memory.

@@

An important thing to note, however, is that unlike the problems most GPU libraries tries to solve we do not attempt so solve large linear equations or invert large matrices; our matrices are small, but we have a very large number of them. Fortunately Nvidia recently released a highly optimized batched linear equation solver tailored for this particular task. It is a Gauss Jordan based with support for partial pivoting and complex numbers. The only downside to using it in our application is that it does not exploit the Hermitian property of our sample covariance matrix. A better choice might be a solver based on Cholesky decomposition. These are designed for Hermitian positive semi-definite matrices, and require roughly half the arithmetic operations compared to a Gauss Jordan solver.

@@

\ifPeerReview\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/code_assessment.eps}
\caption{Execution time of an arithmetic-only and a memory-only version of the MVDR code. A dataset from an $M=32$ array was processed for all $L$ using $K=1$, and the mean execution time for a 1\;Mpx image was used here. From this plot we can infer that the kernel building $\eR$ is memory bound, as the time the kernel spends performing memory transactions is higher than the corresponding time it spends carrying out arithmetical operations. Furthermore, when the total runtime is larger than the restricted kernels this can largely be attributed to latency, which we can see that building $\eR$ suffers from with the chosen parameters.}\label{code_assessment}
\end{figure}
\fi%

@@

The beamformer output $z$ is computed on a per-pixel basis by first computing the MVDR weights $\w$, which are a mere scaled version of $\eRi\1$ (\ref{weights}). Then the sum in (\ref{finalZ}) is found by assigning a group of $L$ threads to respective elements of $x_l$, which then proceed to accumulate these for all $N_L$ subarrays. The resulting data vector is finally multiplied with the weight vector using the same threads, and then a single thread is used to sum these products to obtain $z$.

@@

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske7.eps}
\caption{HISAS sidescan sonar (SSS) image of the shipwreck Holmengraa that lies on a slanted seabed at 77\;m depth outside of Horten, Norway. The image was processed with $M=32$, $L=16$, $K=1$ and $d=1\%$. Prior to display the image was linearly upinterpolated by a factor 2 in azimuth making its total size 1.46\;Mpx. Note how MVDR improves edge definition and reduces noise in shadow regions.}\label{holmengraa}
\end{figure}
\fi%

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske8.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/benchmark.eps}
\fi%
\caption{\protect MVDR benchmarks. A 1\;Mpx image from a $M=32$ channel array was processed for all $L$, and for $K\in\{0,1,2\}$.\newline \emph{Top:} The time the GPU spent on building $\eR$, solving $\eRi\1$, and computing $z$. Note the major speedup of building $\eR$ when compared to the complexity plot in Fig. \ref{mvdr_complexity}.\newline
\emph{Bottom:} Compared to a reference Matlab and single thread C implementation running on a CPU the GPU offered a speedup of 2-3 orders of magnitude, but these numbers are somewhat misleading. If the C implementation was properly optimized we expect the GPU to be no more than a factor 5-10 faster, even if its theoretical peak performance is ~20 times higher than that of the CPU.}\label{benchmarks}
\end{figure}

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske9.eps}
\caption{Execution time of an arithmetic-only and a memory-only version of the MVDR code. A dataset from an $M=32$ array was processed for all $L$ using $K=1$, and the mean execution time for a 1\;Mpx image was used here. From this plot we can infer that the kernel building $\eR$ is memory bound, as the time the kernel spends performing memory transactions is higher than the corresponding time it spends carrying out arithmetical operations. Furthermore, when the total runtime is larger than the restricted kernels this can largely be attributed to latency, which we can see that building $\eR$ suffers from with the chosen parameters.}\label{code_assessment}
\end{figure}
\fi%

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske10.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/code_assess_flops.eps}
\fi
\caption{\protect Code efficiency. An estimate of the number of floating point operations per second (Flop/s), found by dividing the theoretical complexity curves by actual run-times. This is a cruel measure as it does not include any other instructions than the actual arithmetic operations in the MVDR computation.}\label{code_assess_flops}
\end{figure}

@@

We have shown that for systems containing up to 32 channels the problem can be largely mitigated by building the covariance matrix in a clever way, and by making use of the massive computational power available in modern GPUs. We were able to improve upon the run-time of a single thread C-implementation by roughly two orders of magnitude. For most choices of parameters the GPU was able to create images at ~1 Mpx/s at an average data processing rate of ~40\;GFlop/s. This is less than 5\% of the peak performance of the GPU, but we believe it to be near optimal given the constraints of memory bandwidth and the sequential nature of some parts of the MVDR implementation.

@@

% \subsection{Iterative methods}
% 
% \begin{itemize}
% \item Woodbury
% \item Conjugated gradients
% \end{itemize}
% 
% 
% \subsection{Beamspace}
% 

>>>>

\author{Jo~Inge~Buskenes,~\IEEEmembership{Student Member,~IEEE,} %
        Jon~Petter~\AA{}sen,~\IEEEmembership{Student Member,~IEEE,} %
        Carl-Inge~Colombo~Nilsen,~\IEEEmembership{Member,~IEEE,} %
        Andreas~Austeng,~\IEEEmembership{Member,~IEEE}%
\thanks{J. I. Buskenes, C.-I. C. Nilsen and A. Austeng are with the Department of Informatics, University of Oslo, Norway.}%
\thanks{J. P. \AA{}sen is with MI-Lab, Norwegian University of Science and Technology, Trondheim, Norway.}% <-this % stops a space
% \thanks{Manuscript received April 19, 2005; revised January 11, 2007.}
}

@@

In this study we show that for active sonar systems up to 32 channels, the computation time can be significantly reduced by performing arithmetic optimizations, and by implementing the MVDR on a Graphics Processing Unit (GPU). We point out important hardware limitations for these devices, and assess the design in terms of how efficiently it is able to use the GPU's resources. On a quad-core Intel Xeon system with a high-end Nvidia GPU, our GPU implementation renders more than a million pixels per second (1\;MP/s). Compared to our initial CPU implementation the optimizations described herein led to a speedup of more than 2 orders of magnitude, or an expected 5-10 times improvement had the CPU received similar optimization effort. This throughput enables real-time processing of sonar data, and makes the MVDR a viable alternative to conventional methods in practical systems.
% If the HISAS1030 was attached to a platform traveling at 1.8m/s, creating full-coverage sectorscan images in real-time would be possible with this performance.

% We benchmarked optimized implementations of the MV beamformer in MATLAB and C, running on a quad-core Intel Xeon E5620, and in CUDA running on a high-end nVidia card. The beamformers operated on data from the 32x4 element Kongsberg Martime HISAS1030 sonar, using a conservative set of robustification parameters. 
%The CPU would then be free to take on the rest of the beamforming as well.
\end{abstract}

@@

% Robustification:
% R estimate accuracy vs. dynamic trackability: (1) Trade observation time for bandwidth (2) Partial adaptiveness (less degrees of freedom than channels) (3) Quadratic constraint on sensor weights to limit the norm to some preset value.
% -  
To ensure MVDR robustness, the literature suggests combining means such as temporal and spatial averaging, and regularization~\cite{VanTrees2002,Kailath1985}. The complexity issue, on the other hand, can be handled by introducing well-founded approximations. For instance, some studies assume data stationarity which allows the formation of a Toeplitz-structured covariance matrix that is simpler to invert~\cite{Asl2012a,Jakobsson2000}. Other studies perform MVDR beamforming in beamspace (the spatial frequency domain), which can be considered sparse due to the limited angular extent of the received beam~\cite{VanTrees2002,Nilsen2009}. Exploiting this can lead to considerable performance improvements, especially in high channel count systems.

@@

In this work, we show that such approximations may not be necessary for sonar systems up to 32 channels, since here the computation time can be reduced by 1-2 orders of magnitude by implementing MVDR on a Graphics Processing Unit (GPU). Similar work may be found in e.g. medical ultrasound imaging, where Chen \emph{et al.} have demonstrated a GPU implementation of the MVDR that operates on real valued data~\cite{Chen2011,Chen2011a}. While this implementation is fast, the real valued data constraint does not allow the creation of asymmetric and shifted responses, and hence prevents the MVDR's potential to be fully reached. Our implementation is not restricted in this manner.

The implementation is adapted to a well sampled 100\;kHz wideband active sonar system operating in the near field. The system has a range resolution comparable to the sample period so our MVDR implementation runs near single snapshot mode. To deal with this and avoid signal cancellation we compute and average the covariance estimates from overlapping subarrays~\cite{Kailath1985}. This may resemble techniques such as "subarray MVDR"~\cite{Chapman1976}, but unlike that technique we are not using subarrays to reduce complexity but to make MVDR in an active system possible. In related work, we investigated using the same MVDR GPU implementation in cardiac ultrasound imaging~\cite{Asen2012,Asen2013}, and in active sonar imaging we have compared the performance of this implementation to comparable adaptive methods~\cite{Buskenes2013}.

@@

Various adaptive beamformers target this limitation by allowing the weights to change for each pixel to better fit the dynamic nature of the incoming wavefield. In other words, they attempt to use the \emph{a priori} information present in the data to improve image quality. The MVDR beamformer is one such method. It finds the set of complex weights that minimizes the beamformer's expected output power, while ensuring unity gain in the look direction~\cite{Capon1969}. This is a convex optimization problem that can be solved using Lagrange multipliers to yield the solution:
\begin{gather}
\vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
\end{gather}
where $\1$ is a row vector of ones that represents broadside steering, and $\R=E\{\x[n]\x\H[n]\} \in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array. Since we pre-steer our data to every pixel in the image we will always look to broadside, but (\ref{weights}) can be easily generalized by substituting $\1$ with a suitable steering vector. Since $\R$ is unknown, we estimate it by computing a sample covariance matrix $\eR$. In this computation we perform some degree of:
\begin{itemize}
\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985},
\item \emph{temporal averaging} over an interval comparable to the pulse length (1-5 samples) to maintain true speckle statistics~\cite{Synnevag2009}, and
\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
\end{itemize}%
%
% will perform some degree of \emph{spatial averaging} to avoid signal cancellation caused by coherent echoes~\cite{Kailath1985}, \emph{temporal averaging} to maintain true speckle statistics~\cite{Synnevag2009}, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Cox1987,Maksym1979}. 
%
Combined, these steps will also ensure a numerically well conditioned $\eR$.

To form $\eR$ we will first compute an intermediate sample covariance matrix $\breve{\R}$, which we form using temporal and spatial averaging. To do this we need to segment our array into subarrays. If we let $x_l[n]$ represent the data vector from subarray $l$,
\begin{gather}
\x_l[n] = \bmat{x_l[n] & x_{l+1}[n] & \dots & x_{l+L-1}[n]}\T,
\end{gather}
then $\breve{\R}$ can be calculated as:
\begin{gather}
\breve{\R}[n] =  \frac{1}{N_K N_L} \sumb{l=0}{M-L}\sumb{n'=n-K}{n+K} \x_l[n']\x_l\H[n'] \in\mathbb{C}^{L,L},\label{spatialR}
\end{gather}
where $N_K = 2K+1$ is the number of temporal samples to perform averaging over, and $N_L = M-L+1$ is the number of subarrays.

@@

We decided to investigate the feasibility of running MVDR on a GPU by mapping it to a Nvidia GeForce Quadro 6000, which performance-wise is or par with or perhaps slightly inferior to the Geforce 480. This is a high-end Compute Unified Device Architecture (CUDA) enabled GPU based on Nvidia's Fermi architecture, which as now been superseded by the Kepler architecture. The code was written in Nvidia's ``C for CUDA'' framework, which adds a few extra features to the C language to specify how to distribute work across a large number of threads. It would have been possible to use GPUs from AMD and the cross-platform OpenCL framework from the Khronos group instead, but CUDA has a batch linear equation solver~\cite{Nvidia} that no OpenCL library provides. As will be discussed later, this is a key component in our design.

@@

As observed in Fig. \ref{mvdr_complexity}, a direct computation of $\eR$ by implementing (\ref{spatialR}) and (\ref{finalR}) is the greatest computational burden. We target this issue by first \emph{minimizing arithmetic operations}, then aim to get as close to the peak \emph{arithmetic throughput} on the GPU as possible.

\subsubsection{Minimizing Arithmetic Operations}

In Fig. \ref{mvdr_build_R} we illustrate how to reduce the arithmetic complexity of building $\eR$ in a system with $M=5$ sensors, a subarray size of $L=3$ and temporal averaging is set to $K=1$. Here the spatial sum is carried out before the temporal sum. To reduce arithmetic operations we may:
%
\begin{enumerate}
\item exploit the fact that $\eR$ is Hermitian positive semi-definite to compute only one half of it,
\item avoid redundant multiplications, and
\item avoid redundant summations by first computing the upper row of $\eR$ and then add/subtract to these iteratively to find the diagonals of $\eR$.
\end{enumerate}
%
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske5.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/buildR-breakdown.eps}
\fi%
\caption{Arithmetic optimization of computing $\eR$: Relative reduction in arithmetic complexity compared to the initial implementation shown in Fig. \ref{mvdr_complexity} (higher is better) . Note how the arithmetic count is reduced by a factor 4-10 in the memory optimized case, and by a factor 6-22 in the instruction optimized case.}\label{mvdr_complexity_optimized}
\end{figure}

To study the effect of these optimizations we altered the complexity formulas to take them into account. Exploiting symmetry and performing iterative summing (step 1 and 3) is always desirable, but avoiding redundant multiplications (step 2) comes at the cost of increased memory consumption. When all optimizations are applied we call it a "minimum instructions" solution, and when step 2 is left out we call it a "minimum instructions" solution. Fig. \ref{mvdr_complexity_optimized} compares the two solutions, and formulas may be found in appendix \ref{mvdr_formulas}. Here the complexity curves are relative to the reference implementation in Fig. \ref{mvdr_complexity}. Both solutions reduce the complexity considerably, and recomputing the multiplications does not make that much of a difference. We selected the memory minimized version since our algorithm is severely memory bound, as we will see next.

@@

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske6.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/mvdr_implementation.eps}
\fi%
\caption{MVDR implementated on a GPU. We do this in 3 steps, where each step process the full image before moving on to the next step. \emph{Step~1}: The sample covariance matrix $\eR$ is formed by threads running along its diagonals. This allows spatial averaging to be implemented in a computationally efficient manner and minimizes inter-thread communication. \emph{Step~2}: $\eRi\1$ is computed using the heavily optimized batched linear equation solver from Nvidia~\cite{Nvidia}. \emph{Step~3}: The beamformer output $z$ is computed in a straight forward fashion using $L$ threads that first sum the subarrays and then apply the MVDR weighting function. A single thread finally sum all the channels up.}\label{mvdr_implementation}
\end{figure}

@@

When we compute $\eR$, we very rarely perform more than 1-3 floating point operations for every float read or written to memory. Unfortunately, the GPU prefers kernels to be more arithmetically intensive than this. This can be inferred from Table \ref{throughputs}, where the peak bandwidth of the three types of GPU memory are compared to the peak arithmetic throughput (see appendix \ref{throughput} for derivations). Global memory (RAM), in which all data must reside at some point, is at best only able to move one float for every 30 floats processed by the CUDA cores, and potentially much worse if care is not taken to use "coalesced" memory access patterns (that avoid memory bank conflicts). This is why the usage of global memory should be minimized. Shared memory, on the other hand, is a very fast level 1 cache that is shared by all threads on an SM. At peak performance it is almost able to keep up with the computing units, but to obtain this performance we must make sure that:
\begin{enumerate}
\item the data needed to compute a pixel can fit in shared memory, while
\item the access patterns we use promote maximum bandwidth.
\end{enumerate}
 %Further alleviation is also possible by exposing instruction level parallelism within a thread, as this promotes the use of registers with a bandwidth at least 6 times higher than that of shared memory~\cite{Vasilyy}\todo{so why don't we? rephrase}.

@@

In Fig. \ref{mvdr_implementation} we illustrate how we get around these challenges. First we make each SM compute entire range lines of pixels, then load the subset of $\x$ that all these pixels depend upon into shared memory. This lets us perform temporal averaging without having to read additional channel data.

\ifPeerReview\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/plot_holmengraa_L16_Navg1.eps}
\caption{HISAS sidescan sonar (SSS) image of the shipwreck Holmengraa that lies on a slanted seabed at 77\;m depth outside of Horten, Norway. The image was processed with $M=32$, $L=16$, $K=1$ and $d=1\%$. Prior to display the image was linearly upinterpolated by a factor 2 in azimuth making its total size 1.46 megapixels (MP). Note how MVDR improves edge definition and reduces noise in shadow regions.}\label{holmengraa}
\end{figure}
\fi%

Second, we assign $L$ threads per pixel to traverse the diagonals of $\eR$. On the top row a full computation is carried out, then that row is saved back to global memory following a coalesced access pattern that maximizes global memory bandwidth. For subsequent rows the threads move along the diagonals while performing iterative summations; the result from the previous element on the diagonal is updated by adding and removing the correlation coefficients that enters and exits the sum, respectively. To minimize memory consumption, we compute the coefficients again every time we need them. When a thread has finished up a diagonal, we have them wrap around to compute one of the diagonals in the lower triangular of $\eR$. Since $\eR$ is conjugate symmetric, the values in the leftmost column is obtained by a complex conjugate copy of the relevant value in the first row. This is slightly non-optimal as it causes divergent branching, but is only needed because the solver is Gauss Jordan based and thus require a full $\eR$. Combined these steps balance the load evenly on all threads, is almost completely free of arithmetic redundancy, and consumes less memory.

@@

An important thing to note, however, is that unlike the problems most GPU libraries tries to solve we do not attempt so solve large linear equations or invert large matrices; our matrices are small, but we have a very large number of them. Fortunately Nvidia recently released a highly optimized batched linear equation solver tailored for this particular task. It is Gauss Jordan based and supports partial pivoting and complex numbers. The only downside to using it in our application is that it does not exploit the Hermitian property of our sample covariance matrix. A better choice might be a solver based on Cholesky decomposition. These are designed for Hermitian positive semi-definite matrices, and require roughly half the arithmetic operations compared to a Gauss Jordan solver.

@@

\ifPeerReview\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/code_assessment.eps}
\caption{Execution time of an arithmetic-only and a memory-only version of the MVDR code. A dataset from an $M=32$ array was processed for all $L$ using $K=1$, and the mean execution time for a 1\;MP image was used here. From this plot we can infer that the kernel building $\eR$ is memory bound, as the time the kernel spends performing memory transactions is higher than the corresponding time it spends carrying out arithmetical operations. Furthermore, when the total runtime is larger than the restricted kernels this can largely be attributed to latency, which we can see that building $\eR$ suffers from with the chosen parameters.}\label{code_assessment}
\end{figure}
\fi%

@@

The beamformer output $z$ is computed on a per-pixel basis by first computing the MVDR weights $\w$, which are merely scaled versions of $\eRi\1$ (\ref{weights}). Then the sum in (\ref{finalZ}) is found by assigning a group of $L$ threads to respective elements of $x_l$, which then proceed to accumulate these for all $N_L$ subarrays. The resulting data vector is finally multiplied with the weight vector using the same threads, and then a single thread is used to sum these products to obtain $z$.

\subsection{Summary of GPU implementation}

The GPU implementation can be summarized as follows:
\begin{itemize}
\item Pixels are processed in batches.
\item Multiple threads per pixel.
\item $\eR$ is computed directly from raw data to minimize memory footprint.
\item Inversion step is performed by Nvidias batch linear equation solver.
\item Design makes efficient use of cache and data transfers are kept to a minimum.
\end{itemize}

Compared to our CPU implementations, where:
\begin{itemize}
\item Pixels are processed one by one.
\item Single thread per pixel.
\item Computation of $\eR$ is optimized for minumum number of instructions, as in \ref{mvdr_build_R}. Only upper half of $\eR$ is computed (see next point).
\item Inversion step carried out by a custom inline Cholesky based solver that only requires the upper triangular of $eR$.
\item Design has generally not been optimized for efficient use of memory.
\end{itemize}

@@

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske7.eps}
\caption{HISAS sidescan sonar (SSS) image of the shipwreck Holmengraa that lies on a slanted seabed at 77\;m depth outside of Horten, Norway. The image was processed with $M=32$, $L=16$, $K=1$ and $d=1\%$. Prior to display the image was linearly upinterpolated by a factor 2 in azimuth making its total size 1.46\;MP. Note how MVDR improves edge definition and reduces noise in shadow regions.}\label{holmengraa}
\end{figure}
\fi%

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske8.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/benchmark.eps}
\fi%
\caption{\protect MVDR benchmarks. A 1\;MP image from a $M=32$ channel array was processed for all $L$, and for $K\in\{0,1,2\}$.\newline \emph{Top:} The time the GPU spent on building $\eR$, solving $\eRi\1$, and computing $z$. Note the major speedup of building $\eR$ when compared to the complexity plot in Fig. \ref{mvdr_complexity}.\newline
\emph{Bottom:} Compared to a reference MATLAB and single thread C implementation running on a CPU the GPU offered a speedup of 2-3 orders of magnitude, but these numbers are somewhat misleading. If the C implementation was properly optimized we expect the GPU to be no more than a factor 5-10 faster, even if its theoretical peak performance is ~20 times higher than that of the CPU.}\label{benchmarks}
\end{figure}

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske9.eps}
\caption{Execution time of an arithmetic-only and a memory-only version of the MVDR code. A dataset from an $M=32$ array was processed for all $L$ using $K=1$, and the mean execution time for a 1 megapixel (MP) image was used here. From this plot we can infer that the kernel building $\eR$ is memory bound, as the time the kernel spends performing memory transactions is higher than the corresponding time it spends carrying out arithmetical operations. Furthermore, when the total runtime is larger than the restricted kernels this can largely be attributed to latency, which we can see that building $\eR$ suffers from with the chosen parameters.}\label{code_assessment}
\end{figure}
\fi%

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske10.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/code_assess_flops.eps}
\fi
\caption{\protect Code efficiency. An estimate of the number of floating point operations per second (Flop/s), found by dividing the theoretical complexity curves by actual run-times. This is a crude measure as it does not include any other instructions than the actual arithmetic operations in the MVDR computation.}\label{code_assess_flops}
\end{figure}

@@

We have shown that for systems containing up to 32 channels the problem can be largely mitigated by building the covariance matrix in a clever way, and by making use of the massive computational power available in modern GPUs. We were able to improve upon the run-time of a single thread C-implementation by roughly two orders of magnitude. For most choices of parameters the GPU was able to create images at $\sim$1\;MP/s at an average data processing rate of ~40\;GFlop/s. This is less than 5\% of the peak performance of the GPU, but we believe it to be near optimal given the constraints of memory bandwidth and the sequential nature of some parts of the MVDR implementation.

@@
